\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-05-05 analysis/paper/paper.tex
edit jesse's section

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras; the interested
reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
Saarela \citeyearpar{saarela2016case}. First, let
\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
event of interest for individual \(i=1, \ldots,n\). For simplicity, we
will consider Type I censoring due to the end of follow-up at time
\(\tau\) (the general case of non-informative censoring is treated in
Saarela \citeyearpar{saarela2016case}). We assume a continuous time
model, which implies that the counting process jumps are less than or
equal to one. We are interested in modeling the hazard functions
\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]

@@

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual contributions of the form
\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, we can define an estimating equation for \(\theta\) as
follows:
\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
When a logarithmic link function is used for modeling the hazard
function, the above expression is of a logistic regression form with an
offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
in the case-base sampling mechanism are person-moments, rather than
individuals, and the parameters to be estimated are hazards or hazard
ratios rather than odds or odds ratios. Generally, an individual can
contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the logarithm of this
estimating function has mean zero at the true value \(\theta=\theta_0\),
and that the resulting estimator \(\hat{\theta}\) is asymptotically
normally distributed.

@@

The sampling scheme selected for \code{sampleCaseBase} was described in
Hanley \& Miettinen \citeyearpar{hanley2009fitting}: we first sample
along the ``person'' axis, proportional to each individual's total
follow-up time, and then we sample a moment uniformly over their
follow-up time. This sampling scheme is equivalent to the following
picture: imagine representing the total follow-up time of all
individuals in the study along a single dimension, where the follow-up
time of the next individual would start exactly when the follow-up time
of the previous individual ends. Then the base series could be sampled
uniformly from this one-dimensional representation of the overall
follow-up time. In any case, the output is a dataset of the same class
as the input, where each row corresponds to a person-moment. The
covariate profile for each such person-moment is retained, and an offset
term is added to the dataset. This output could then be used to fit a
smooth hazard function, or for visualization of the base series.

@@

\hypertarget{case-study-1european-randomized-study-of-prostate-cancer-screening}{%
\section{Case study 1--European Randomized Study of Prostate Cancer
Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}}

@@

\hypertarget{case-study-2bone-marrow-transplant}{%
\section{Case study 2--Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}}

@@

\hypertarget{case-study-3support-data}{%
\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. In the next two case studies, we explore
more advanced features.

For the third case study, we will look at regularized estimation for the
hazard function. This can be performed using a combination of case-base
sampling and regularized logistic regression. To demonstrate this
capability, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within 5 hospitals in the
United States of America. Imputation was conducted using the \pkg{mice}
package in \proglang{R} with default settings \citep{mice}. Before
imputation, there were a total of 9105 individuals and 47 variables. Of
those variables, a few were removed before imputation. We removed three
response variables: hospital charges, patient ratio of costs to charges
and patient micro-costs. Next, we removed hospital death as it was
directly informative of our event of interest, death. We also removed
functional disability and income as they are ordinal covariates, because
ordinal variables are treated as factors in our current model. Finally,
we removed 8 covariates relating to the results of previous findings: we
removed SUPPORT day 3 physiology score (sps), APACHE III day 3
physiology score (aps), SUPPORT model 2-month survival estimate, SUPPORT
model 6-month survival estimate, Physician's 2-month survival estimate
for pt., Physician's 6-month survival estimate for pt., Patient had Do
Not Resuscitate (DNR) order and Day of DNR order (\textless{}0 if before
study). Of these, sps and aps were added on after imputation, as they
were missing only 1 observation and permitted us to compare findings
between our model and their's.

First we impute manually using recommended normal values for
physiological measures \citep{harrell_2020}. Next, we impute a single
dataset using \pkg{mice} with default settings. After imputation, we
noted that the covariate for surrogate activities of daily living was
not imputed. This is due to colinearity between the other two covariates
for activities of daily living. As such, surrogate activites of daily
living was removed from our model. After these imputations, we have a
total of 9104 observations and 30 variables, two of which are sps and
aps, which will work as a baseline for our comparison to previous
findings. This dataset is available Its worth noting that these scores
were developed for a period of 180 days, which makes our comparison over
5.56 years exploratory. Also, our single imputation introduces bias in
the data. As the goal here is purely demonstrative, this bias will be
consistent with any of the following models. As such, their comparison
is still valid, though any interpretation requires an awareness of said
bias. A description of used variables can be found in Table
\ref{tab:support1} and a breakdown of each categorical variable in Table
\ref{tab:support2}. The original data was obtained from the Department
of Biostatistics at Vanderbilt University \citep{harrell_2020}. The
imputed dataset is available as part of the \pkg{casebase} package.

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}. As we can see, the incidence density is not
constant.

{\centering \includegraphics{../figures/support-poptime-1}

\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
\end{figure}
\end{CodeChunk}

Before generating an absolute risk curve using regularized logistic
regression, we will explore three options as a baseline comparison,
beyond a Kaplan-meier curve. The original study compared these scores
over a span of 180 days, which may not be comparable over the full 5.56
years. Therefore, We will examine three cox regression scenarios: sps
only, aps only and all covariates excluding sps and aps. Next, we will
compare each modeling using the Brier score \citep{graf1999ass}. The
Brier score is able to demonstrate both discrimination and calibration
in a single model \citep{graf1999ass} and provide a comparable metric
between semi-parametric and parametric methods in the survival context.
For all the following analyses in this case study, we will use 80\% of
the observations to train our models, while 20\% will be used to
generate absolute risk curves, all of which will be averaged to
demonstrate each model's predictive ability. Over 5.56 years,
\ref{fig:baselines} demonstrates that all the cox models have similar,
if not identical average predictions. Figure \ref{fig:brier1} examines
the brier scores as they change over time. As the full model has the
lowest brier score on overall, it will be used in the following
comparison to regularized cox, regularized logistic regression and the
unadjusted Kaplan-Meier curve.

@@

\caption{\label{fig:baselines} AbsoluteRisk curves from hazard models. Pink: All covariates without SPS nor APS. Green: APS only model. Blue: SPS only model. Purple: unadjusted curve. All three models underestimate the risk in comparison to the unadjusted curve.}\label{fig:baselines}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:brier1} Brier scores over time. Full model (Cox Full) performs best, while the other two (APS only and SPS only) are comparable.}\label{fig:brierpt1}
\end{figure}
\end{CodeChunk}

To get the absolute risk following regularized cox regression, this
required that we work around limitations in existing packages. To make
use of the \pkg{survival} package, we first get our hazard coefficients
using \pkg{glmnet}'s coxnet implementation for lasso penalization
\citep{regpathcox}. Next, we fit a cox model using the survival package
with the selected variables only, from our previous fit. Finally, we
replaced the coefficients in the survival object with those from our
glmnet fit, allowing us to make use of the absolute risk tools available
in the \pkg{survival} package.

For our penalized logistic regression model, we opted for a logistic
function of time, resulting in a Weibull distribution assumption. As
casebase makes use of the glmnet package, we interact with the
fitsmoothhazard.fit function using a matrix interface, where y contains
the time and event variables and x contains all other variables we would
like to include in the model. For the SUPPORT dataset, all factor
variables were converted into dummy variables. Here, we used lasso
penalization by setting alpha to 1, leading to our smooth hazard
function. We then take our hazard and use it to predict the absolute
risk of each individual in our test set, over time. The probabilites
over time for each individual are averaged, resulting in the smooth
absolute risk curve in figure \ref{fig:cs3Final}. On average, the Brier
score is lower in our penalized logistic regression model, compared to
the other models (fig.~\ref{fig:cs3FinalBrier}).

This case study serves to demonstrate how regularization can be used in
a survival context, using logistic regression. Our approach performs
comparably to Cox regression, demonstrating its validity in an applied
setting.

@@

\begin{CodeInput}
R> 
R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.facto=c(0,rep(1,ncol(x))))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Estimating the absolute risk curve using the newData parameter.
R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
\end{CodeInput}
\end{CodeChunk}

@@

{\centering \includegraphics{../figures/unnamed-chunk-2-1}

@@

\caption{\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (pink), penalized Cox (green), penalized logistic (blue) and Kaplan-Meier (purple).}\label{fig:unnamed-chunk-2}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between a null model (Kaplan-Meier, black), Cox regression (yellow), penalized Cox regression (blue) and penalized logistic regression (green).}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

\caption{\label{fig:cs3FinalBrier} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (blue), penalized Cox (Green), penalized logisitic (red). If a coefficient is estimated to be 0, it is removed from the plot}\label{fig:coefplots}
\end{figure}
\end{CodeChunk}

\hypertarget{case-study-4stanford-heart-transplant-data}{%
\section{Case study 4--Stanford Heart Transplant
Data}\label{case-study-4stanford-heart-transplant-data}}

@@

This report was generated on 2020-04-09 13:28:46 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 3.6.3 (2020-02-29)
#> Platform: x86_64-w64-mingw32/x64 (64-bit)
#> Running under: Windows 10 x64 (build 18363)
#> 
#> Matrix products: default
#> 
#> locale:
#> [1] LC_COLLATE=English_United States.1252 
#> [2] LC_CTYPE=English_United States.1252   
#> [3] LC_MONETARY=English_United States.1252
#> [4] LC_NUMERIC=C                          
#> [5] LC_TIME=English_United States.1252    
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.5.5               dotwhisker_0.5.0         
#>  [3] prodlim_2019.11.13        riskRegression_2020.03.27
#>  [5] pracma_2.2.9              kableExtra_1.1.0         
#>  [7] data.table_1.12.8         glmnet_3.0-2             
#>  [9] Matrix_1.2-18             casebase_0.2.1.9001      
#> [11] survival_3.1-11           magrittr_1.5             
#> [13] forcats_0.5.0             stringr_1.4.0            
#> [15] dplyr_0.8.5               purrr_0.3.3              
#> [17] readr_1.3.1               tidyr_1.0.2              
#> [19] tibble_3.0.0              ggplot2_3.3.0            
#> [21] tidyverse_1.3.0          
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-2          colorspace_1.4-1   
#>  [4] ellipsis_0.3.0      ggstance_0.3.4      htmlTable_1.13.3   
#>  [7] base64enc_0.1-3     fs_1.4.1            rstudioapi_0.11    
#> [10] farver_2.0.3        MatrixModels_0.4-1  fansi_0.4.1        
#> [13] mvtnorm_1.1-0       lubridate_1.7.8     xml2_1.3.1         
#> [16] codetools_0.2-16    knitr_1.28          Formula_1.2-3      
#> [19] jsonlite_1.6.1      cluster_2.1.0       dbplyr_1.4.2       
#> [22] png_0.1-7           compiler_3.6.3      httr_1.4.1         
#> [25] backports_1.1.6     assertthat_0.2.1    cli_2.0.2          
#> [28] acepack_1.4.1       htmltools_0.4.0     quantreg_5.55      
#> [31] tools_3.6.3         gtable_0.3.0        glue_1.4.0         
#> [34] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.2.4        
#> [37] nlme_3.1-145        iterators_1.0.12    xfun_0.12          
#> [40] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
#> [43] MASS_7.3-51.5       zoo_1.8-7           scales_1.1.0       
#> [46] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [49] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
#> [52] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
#> [55] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [58] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [61] mets_1.2.7.1        rlang_0.4.5         pkgconfig_2.0.3    
#> [64] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
#> [67] labeling_0.3        cmprsk_2.2-9        tidyselect_1.0.0   
#> [70] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
#> [73] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
#> [76] pillar_1.4.3        haven_2.2.0         foreign_0.8-76     
#> [79] withr_2.1.2         mgcv_1.8-31         nnet_7.3-13        
#> [82] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
#> [85] timereg_1.9.4       jpeg_0.1-8.1        grid_3.6.3         
#> [88] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_3.6.3       
#> [94] munsell_0.5.0       viridisLite_0.3.0
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    master C:/Users/jesse/Documents/gits/cbpaper
#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
#> Head:     [b31e195] 2020-03-31: Fixed penal. on time and Brier
\end{CodeOutput}
\end{CodeChunk}

>>>>

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras; the interested
reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
Saarela \citeyearpar{saarela2016case}. First, let
\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
event of interest for individual \(i=1, \ldots,n\). For simplicity, we
will consider Type I censoring due to the end of follow-up at time
\(\tau\) (the general case of non-informative censoring is treated in
Saarela \citeyearpar{saarela2016case}). We assume a continuous time
model, which implies that the counting process jumps are less than or
equal to one. We are interested in modelling the hazard functions
\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]

@@

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual contributions of the form
\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, we can define an estimating equation for \(\theta\) as
follows:
\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
When a logarithmic link function is used for modelling the hazard
function, the above expression is of a logistic regression form with an
offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
in the case-base sampling mechanism are person-moments, rather than
individuals, and the parameters to be estimated are hazards or hazard
ratios rather than odds or odds ratios. Generally, an individual can
contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the logarithm of this
estimating function has mean zero at the true value \(\theta=\theta_0\),
and that the resulting estimator \(\hat{\theta}\) is asymptotically
normally distributed.

@@

The sampling scheme selected for \code{sampleCaseBase} was described in
Hanley \& Miettinen \citeyearpar{hanley2009fitting}: we first sample
along the ``person'\,' axis, proportional to each individual's total
follow-up time, and then we sample a moment uniformly over their
follow-up time. This sampling scheme is equivalent to the following
picture: imagine representing the total follow-up time of all
individuals in the study along a single dimension, where the follow-up
time of the next individual would start exactly when the follow-up time
of the previous individual ends. Then the base series could be sampled
uniformly from this one-dimensional representation of the overall
follow-up time. In any case, the output is a dataset of the same class
as the input, where each row corresponds to a person-moment. The
covariate profile for each such person-moment is retained, and an offset
term is added to the dataset. This output could then be used to fit a
smooth hazard function, or for visualization of the base series.

@@

\hypertarget{case-study-1european-randomized-study-of-prostate-cancer-screening}{%
\section{Case study 1---European Randomized Study of Prostate Cancer
Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}}

@@

\hypertarget{case-study-2bone-marrow-transplant}{%
\section{Case study 2---Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}}

@@

\hypertarget{case-study-3support-data}{%
\section{Case study 3---SUPPORT Data}\label{case-study-3support-data}}

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. For the third case study, we show how
\pkg{casebase} can also be used for variable selection through
regularized estimation of the hazard function. To do this, we simply
replace logistic regression with a penalized counterpart. In
\pkg{casebase}, we achieve this by using the \pkg{glmnet} package
\citep{friedman2010jss}.

To illustrate this functionality, we use the dataset from the Study to
Understand Prognoses Preferences Outcomes and Risks of Treatment
(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
five American hospitals within individuals who are considered seriously
ill.~The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \pkg{casebase} package. For more information
about this dataset, the reader is encouraged to look at the
documentation in our package. A description of the variables can be
found in Table \ref{tab:support1}, and a breakdown of each categorical
variable appears in Table \ref{tab:support2}.

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{Description of each level within each categorical variable in the SUPPORT dataset.}
\label{tab:support2}
\end{table}

In the original study, the authors developed two scores for predicting
the outcome. These scores are available as part of the data: SUPPORT day
3 physiology score (\code{sps}) and APACHE III day 3 physiology score
(\code{aps}). Using these scores, we will explore three Cox regression
scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only,
\code{aps} only, and all covariates excluding \code{sps} and \code{aps}.
We will compare the three models using a time-dependent version of the
classical Brier score that is adjusted for censoring
\citep{graf1999ass}. The Brier score can be used to assess both
discrimination and calibration in a single model. Moreover, it provides
a metric that can be used to compare parametric and semi-parametric
survival models. To account for overfitting, we will use 95\% of the
observations to train our models, while the remaining 5\% will be used
to compute absolute risk curves and the Brier scores.

\begin{CodeInput}
R> data(support)
R> # Change time to years
R> support$d.time <- support$d.time/365.25
R> 
R> # Split into test and train
R> train_index <- sample(nrow(support), 0.95*nrow(support))
R> test_index <- setdiff(1:nrow(support), train_index)
R> 
R> train <- support[train_index,]
R> test <- support[test_index,]
\end{CodeInput}
\end{CodeChunk}

In Figure\ref{fig:baselines}, we can see that all Cox models have
similar average survival curves. On the other hand, Figure
\ref{fig:brier1} shows that the full model has the lowest Brier score
overall. For this reason, we will use this model in our comparison
between regularized Cox regression, Kaplan-Meier estimation, and the
approach from \pkg{casebase}.

@@

\caption{\label{fig:baselines} Cumulative Incidence curves estimated using Cox regression and Kaplan-Meier (K-M).}\label{fig:baselines}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:brier1} Time-dependent Brier scores.}\label{fig:brierpt1}
\end{figure}
\end{CodeChunk}

For our penalized logistic regression model, we opt for the natural log
of time; recall that this corresponds to a Weibull distribution. For
fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which
is a matrix interface to the function \code{fitSmoothHazard}. In order
to use this interfact, we first create a matrix \code{y} containing the
time and event variables, and a matrix \code{x} containing all remaining
covariates we would like to include in the model space. We used lasso
penalization by setting \code{alpha = 1}. Moreover, we use the argument
\code{penalty.factor} to force the time variable into the final model.

@@

\begin{CodeInput}
R> x <- model.matrix(death ~ . - d.time - aps - sps, 
R+                   data = train)[, -c(1)] # Remove intercept
R> y <- data.matrix(subset(train, select = c(d.time, death)))
R> 
R> # Regularized logistic regression to estimate hazard
R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
R+   family = "glmnet",
R+   time = "d.time", event = "death",
R+   formula_time = ~ log(d.time), alpha = 1,
R+   ratio = 10, standardize = TRUE,
R+   penalty.factor = c(0, rep(1, ncol(x)))
R+ )
\end{CodeInput}
\end{CodeChunk}

Next, we can use the object \code{cbFull} to estimate the absolute risk
curve on the test data. We first create a matrix \code{newx} containing
the relevant variables. Next, using \code{times}, we specify the time
points at which the cumulative incidence will be computed. Finally,
using the parameter \code{s = "lambda.1se"}, we specify at which value
of the tuning parameter we want to retrieve the coefficient estimates.

@@

\begin{CodeInput}
R> # Estimating the absolute risk curve using the newdata parameter
R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
R+                   data = test)[, -c(1)]
R> times <- sort(unique(test$d.time))
R> 
R> abCbFull <- casebase::absoluteRisk(cbFull,
R+   time = times,
R+   newdata = newx,
R+   s = "lambda.1se",
R+   method = "numerical"
R+ )
\end{CodeInput}
\end{CodeChunk}

Our first comparison plot examines the covariates that were selected
between all three hazard functions. Note that Cox regression without
penalization selected more covariates than the other two, but only the
intersection of coefficients with values greater than 0 are displayed in
\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
larger estimated coefficients. This is expected, as there are no
restrictions on the size of coefficients. The smallest coefficient
estimates are found in penalized logistic model, aside from cancer state
(cano).

@@

@@

\caption{\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models. If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
\end{figure}
\end{CodeChunk}

These values do not inform us of model performance. To do so, we take
our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
Figure \ref{fig:cs3Final}. There is separation between the curves, but
it is minimal. To better demonstrate the differences in performance
between the models, we compare Brier scores over time. In Figure
\ref{fig:cs3FinalBrier}, the Brier score is lower in our penalized
logistic regression model compared to both Cox regression models.
However, this difference is minimal. On the other hand, all three models
outperform the Kaplan-Meier model.

{\centering \includegraphics{../figures/cs3Final-1}

\caption{\label{fig:cs3Final} Four Cumulative Incidence curves.}\label{fig:cs3Final}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:cs3FinalBrier} Four curves comparing the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

\hypertarget{case-study-4stanford-heart-transplant-data}{%
\section{Case study 4---Stanford Heart Transplant
Data}\label{case-study-4stanford-heart-transplant-data}}

@@

This report was generated on 2020-05-05 17:12:52 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.0 (2020-04-24)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.4
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.5.6               dotwhisker_0.5.0         
#>  [3] colorspace_1.4-1          prodlim_2019.11.13       
#>  [5] riskRegression_2020.02.05 pracma_2.2.9             
#>  [7] kableExtra_1.1.0          data.table_1.12.8        
#>  [9] glmnet_3.0-2              Matrix_1.2-18            
#> [11] casebase_0.2.1.9001       survival_3.1-12          
#> [13] magrittr_1.5              forcats_0.5.0            
#> [15] stringr_1.4.0             dplyr_0.8.5              
#> [17] purrr_0.3.4               readr_1.3.1              
#> [19] tidyr_1.0.2               tibble_3.0.1             
#> [21] ggplot2_3.3.0             tidyverse_1.3.0          
#> 
#> loaded via a namespace (and not attached):
#>  [1] nlme_3.1-147        fs_1.4.1            cmprsk_2.2-9       
#>  [4] lubridate_1.7.8     rms_5.1-4           webshot_0.5.2      
#>  [7] RColorBrewer_1.1-2  httr_1.4.1          numDeriv_2016.8-1.1
#> [10] tools_4.0.0         backports_1.1.6     R6_2.4.1           
#> [13] rpart_4.1-15        Hmisc_4.4-0         DBI_1.1.0          
#> [16] mgcv_1.8-31         nnet_7.3-13         withr_2.2.0        
#> [19] gridExtra_2.3       tidyselect_1.0.0    compiler_4.0.0     
#> [22] quantreg_5.55       cli_2.0.2           rvest_0.3.5        
#> [25] htmlTable_1.13.3    SparseM_1.78        xml2_1.3.2         
#> [28] sandwich_2.5-1      labeling_0.3        checkmate_2.0.0    
#> [31] scales_1.1.0        mvtnorm_1.1-0       polspline_1.1.17   
#> [34] digest_0.6.25       foreign_0.8-78      rmarkdown_2.1      
#> [37] base64enc_0.1-3     rticles_0.14        jpeg_0.1-8.1       
#> [40] pkgconfig_2.0.3     htmltools_0.4.0     dbplyr_1.4.3       
#> [43] htmlwidgets_1.5.1   rlang_0.4.5         readxl_1.3.1       
#> [46] rstudioapi_0.11     VGAM_1.1-2          farver_2.0.3       
#> [49] shape_1.4.4         generics_0.0.2      zoo_1.8-7          
#> [52] jsonlite_1.6.1      acepack_1.4.1       Formula_1.2-3      
#> [55] Rcpp_1.0.4.6        munsell_0.5.0       fansi_0.4.1        
#> [58] lifecycle_0.2.0     multcomp_1.4-13     stringi_1.4.6      
#> [61] yaml_2.2.1          MASS_7.3-51.5       plyr_1.8.6         
#> [64] ggstance_0.3.4      grid_4.0.0          crayon_1.3.4       
#> [67] lattice_0.20-41     haven_2.2.0         hms_0.5.3          
#> [70] timereg_1.9.4       knitr_1.28          pillar_1.4.3       
#> [73] codetools_0.2-16    stats4_4.0.0        reprex_0.3.0       
#> [76] glue_1.4.0          evaluate_0.14       latticeExtra_0.6-29
#> [79] modelr_0.1.6        vctrs_0.2.4         png_0.1-7          
#> [82] foreach_1.5.0       MatrixModels_0.4-1  cellranger_1.1.0   
#> [85] gtable_0.3.0        assertthat_0.2.1    xfun_0.13          
#> [88] viridisLite_0.3.0   iterators_1.0.12    cluster_2.1.0      
#> [91] lava_1.6.7          TH.data_1.0-10      ellipsis_0.3.0
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
#> Head:     [70115f2] 2020-04-22: Merge pull request #20 from Jesse-Islam/master
\end{CodeOutput}
\end{CodeChunk}
